---
title: "CACAO fermentation"
#date: "2020/02/29"
output: 
  revealjs::revealjs_presentation:
    incremental: false
    css: ["css/fonts.css"]
    #css: ["default", "css/sfah.css", "css/fonts.css", "css/ninjutsu.css"]
    self_contained: false
    reveal_plugins: ["notes", "menu", "chalkboard", "zoom"]
    theme: night
    transition: fade
    highlight: pygments
    center: true
    reveal_options:
      chalkboard:
        theme: chalkboard
        toggleNotesButton: true
      slideNumber: true
      previewLinks: false
      slide_level: 1
      #center: true
      mathjax: false
      width: "100%"
      height: "100%"
      margin: 0.2
      fragments: true
      navigationMode: grid
#      minScale: 1
#      maxScale: 1
# Transitions: none - fade - slide - convex - concave - zoom
# Themes: black (default) - white - league - sky - beige - simple serif - blood - night - moon - solarized
# Slide transitions: "default", "fade", "slide", "convex", "concave", "zoom" or "none".

# data-transition-speed controls text

---

```{r setup, include=FALSE}
remove(list = ls())
library(treemap)
library(d3treeR)
library(DiagrammeR)
library(htmlwidgets)
library(DiagrammeRsvg)
library(rsvg)
library(RColorBrewer)
library(collapsibleTree)
library(htmltools)

#Full palette
tree_pal <- c("#56B4E9", "#D55E00", "#009E73", "#CC79A7", "#B6DBFF", "#7F7F7F", "#0072B2")

```


#  <font color="#F5F5F5">overview </font>{ data-background-image="images/paper-bag.jpg" data-background-size="110%"  background-transition-speed="slow" style="font-size: 0em;"}


## <font color="#F5F5F5">outline </font>{ data-background-image="images/collage.png" data-background-size="90%"  background-transition-speed="slow" style="font-size: 0em;"} 

PUT SMALL PICS OF TOPICS

# Tree to Bean to Bar<br/><small>the Process of Making Chocolate</small>


## <font color="#414141">Making cocoa: a complex & multistep process </font> {data-transition="fade-in" data-background="#FFFFFF"}

```{r, workflow, echo = FALSE, fig.align='left', fig.height= 5}
dag <- grViz ("
digraph boxes_and_circles {
  graph [layout = dot, rankdir = LR, overlap = FALSE, outputMode = nodesfirst]

  node [fontname = sans, 
        fontsize = 200, 
        penwidth = 30,
        shape = rectangle,
        style = rounded,
        height = 4.5, 
        width = 6,
        color = '#0072B2',
        margin = '0.91, 0.055']
  0[label = 'Farming'];
  1[label = 'Pod \n storage'];
  5[label = 'Drying'];
  7[label = 'Cleaning'];
  8[label = 'Roasting'];
  10[label = 'Winnowing', URL = 'https://en.wikipedia.org/wiki/Winnowing'];
  12[label = 'Alkanization', URL = 'https://en.wikipedia.org/wiki/Dutch_process_chocolate'];
  14[label = 'Blending'];
  15[label = 'Milling'];
  17[label = 'Pressing'];
  22[label = 'Conching', URL = 'https://en.wikipedia.org/wiki/Conche'];

  node [fontname = sans, 
        fontsize = 200, 
        penwidth = 30,
        shape = rectangle,
        style = rounded,
        height = 4.5, 
        width = 6, 
        color = '#CC79A7', 
        style='filled, 
        rounded', 
        shape = circle]
  4[label = 'Precursor \n formation'];
  6[label = 'Flavor'];
  9[label = 'Flavor'];
  13[label = 'Flavor & \n Color change'];
  24[label = 'Flavor & \n texture '];

  node [fontname = sans, 
        fontsize = 200, 
        penwidth = 30,
        shape = circle,
        style = 'filled, rounded',
        color = '#E69F00']
  11[label = 'Cacao \n nibs'];
  18[label = 'Cocoa  \n Butter'];
  19[label = 'Cocoa  \n Powder'];
  20[label = 'Cocoa  \n beverages'];
  21[label = 'Other \n products'];
  
  node [fontname = sans,
  fontsize = 300]
  2[label = 'Cacao \n Beans', color = '#0000001A', style='filled, rounded', shape = oval, height = 7, width = 12];
  3[label = 'Fermentation', color = '#F0E442', style='filled, rounded', shape = oval, height = 13, margin = '0.91, 0.055'];
  16[label = 'Cocoa  \n  liquor  \n & \n  cocoa \n mass', color = '#E69F00', style='filled, rounded', shape = oval, height = 5, width = 2, fontsize = 200];
  23[label = 'Chocolate', color = '#924900', style='filled, rounded', shape = rectangle, height = 11, width = 14, fontcolor = '#FFFFFF'];

  edge[penwidth=15, 
       color=black, 
       arrowhead = tee, 
       arrowsize = 10];
  
0->1;
1->2[minlen = 3.5]; 
2->3[arrowhead = inv, arrowsize = 6]; 
3->4[arrowhead = dot, arrowsize = 10, minlen = 4.5]; 
3->5; 
5->6[arrowhead = dot, arrowsize = 10, minlen = 4.5]; 
5->7; 
7->8; 
8->9[arrowhead = dot, arrowsize = 10, minlen = 4.5];
8->10; 
11->10[arrowhead = dot, arrowsize = 6, dir = back, minlen = 3.5]; 
10->12[headport = w]; 
12->13[arrowhead = dot, arrowsize = 10, minlen = 4.5]; 
12->14; 
14->15; 
15->16[arrowhead = inv, arrowsize = 6]; 
16->{17 22}[headport = w]; 
17->18[arrowhead = inv, arrowsize = 6]; 
17->19[arrowhead = inv, arrowsize = 6]; 
19->{20 21}[arrowhead = inv, arrowsize = 6]; 
22->23;
22->24[arrowhead = dot, arrowsize = 10, minlen = 3.5];

	graph [nodesep = 2.5, ranksep = 1.25]
{ rank=same; 3, 4 }
{ rank=same; 5, 6 }
{ rank=same; 8, 9 }
{ rank=same; 10, 11 }
{ rank=same; 12, 13 }
{ rank=same; 22, 24 }
{ rank=max; 23 }

}
",  width = 1100)

## This is all for exporting SVG###
#library(DiagrammeRsvg)
#svg <- export_svg(dag)
#writeLines(svg, "graph.svg")
#cat(readLines("graph.svg"), sep = "\n")
#writeLines(svg, "graph.svg")
##################

## This is all for exporting PNG###
#svg <- export_svg(dag) %>%
#  charToRaw() %>%
#  rsvg() %>%
#  png::writePNG("dag.png")

dag

```


##  <font color="#F5F5F5">cacao to cocoa </font>{ data-background-image="images/cacao_to_cocoa.png" data-background-size="85%"  background-transition-speed="slow" style="font-size: 0em;"}


## <font color="#F5F5F5">Volatile compound profiles in beans vs chocolates </font>{ data-background-image="images/sankey.png" background-transition="slide"  data-background-size="95%" style="font-size: 0em;"}


##  <font color="#414141" font-weight="800"> more than 600 volatile compounds in cocoa flavor</font> {background-transition="slide" data-background-image="images/odor_volatiles.png" data-background-size="85%" background-color="#F5F5F5"  style="font-size: 0.8em;"}

# Fermentation


## <font color="#1b1b1b">Fermentation defined</font> {data-background-image="images/fermentation_general.png" data-background-size="85%" style="font-size:0em"}


## <font color="#1b1b1b">history</font> {data-background-image="images/fermentation-history.png" data-background-size="80%" style="font-size:0em"}


## Fermentation in Cacao

<br/>
What is the process of fermentation in cacao?

How does fermentation contribute to the flavor of chocolate?

What factors influence the fermentation process?


# The Cacao Bean

## <font color="#1b1b1b">The Seed</font> {data-background-image="images/beans_unfermented.png" data-background-size="90%" style="font-size:0em"}

## <font color="#1b1b1b">Cacao Pulp <br/> <small>a rich medium for microbial growth</small></font> {data-background-image="images/bean_in_pod.jpg" data-background-size="100%" style="font-size:1.3em"}


```{r setup_treemap, echo=FALSE, include=FALSE}
# dataset
group <- c(
           rep("Sugars",3),
           rep("Plant cell wall",4),
           rep("Citrate",1),
           rep("Protein, peptides, amino acids",4),
           rep("Fat",1),
           rep("Trace",6)
           )
subgroup <- c(
              "Sucrose","Glucose","Fructose",
              "Cellulose","Hemicellulose","Pectin","Lignin", 
              "Citrate", 
              "Total nitrogen", "Free amino acids", "Protein & peptides", "Ammonia",
              "Fat",
              "Vitamins","Potassium", "Calcium", "Iron", "Magnesium", "other"
)
value <- c(
           4.35,3,3.8, #Sugars
           0.9,0.5,1.15,0.26, #Plant cell wall
           1.31, # Citrate
           0.11, 0.15, 0.57, 0.02, # Protein, peptides, amino acids
           0.45, # FAT
           0.05, 0.21, 0.013, 0.004, 0.014, 0.0012 # TRACE
)
data <- data.frame(group,subgroup,value)
 
# basic treemap
p <- treemap(data,
            index=c("group","subgroup"),
            vSize="value",
            vColor = "group",
            type="index",
            fontsize.labels=c(40, 8),
            fontface.labels=c(2,1),
            bg.labels=c("transparent"), 
            palette = tree_pal,
            overlap.labels=0.5,  
            algorithm = "pivotSize",
            title="Pulp composition",
            align.labels=list(c("center", "center"), c("right", "bottom"))  
          )

#friend_pal
#display.brewer.all(colorblindFriendly = TRUE)
#display.brewer.pal(n = 81, name = 'RdBu')
```

```{r style_widget, echo=FALSE, include=FALSE}
style_widget <- function(hw=NULL, style="", addl_selector="") {
  stopifnot(!is.null(hw), inherits(hw, "htmlwidget"))

  # use current id of htmlwidget if already specified
  elementId <- hw$elementId
  if(is.null(elementId)) {
    # borrow htmlwidgets unique id creator
    elementId <- sprintf(
      'htmlwidget-%s',
      htmlwidgets:::createWidgetId()
    )
    hw$elementId <- elementId
  }

  htmlwidgets::prependContent(
    hw,
    htmltools::tags$style(
      sprintf(
        "#%s %s {%s}",
        elementId,
        addl_selector,
        style
      )
    )
  )
}
```


## The Pulp


```{r make_treemap, echo=FALSE, fig.align='right'}
## UNCOMMENT TO REGENERATE STAND-ALONE HTML VERSION
# make it interactive ("rootname" becomes the title of the plot):
# inter <- d3tree(p,  rootname = "Pulp composition", width = "100%")
# inter
# d3tree(p,  rootname = "Pulp composition", width = "100%")
# save the widget
# saveWidget(inter, file="interactiveTreemap.html", selfcontained = TRUE, background = "#111111")

style_widget(
  d3tree(p, rootname = "Pulp composition", width = "100%"),
  addl_selector="text",
  style="font-family:sans; font-size:20px;"
)

## Add the following line OUTSIDE CHUNK TO USE STAND-ALONE HTML VERSION instead
# <iframe src="interactiveTreemap.html" style="border:none;height:1400px;width:1400px"></iframe>

```


## <font color="#1b1b1b">parenchyma</font> {data-background-image="images/parenchyma.png" data-background-size="70%" style="font-size:0em"}

# Cacao Fermentation  <br/> <small>the key step in the formation of aroma precursors</small>

## Why is fermentation important?

::: {style="color: #F5F5F5;"}
* Unfermented beans do not produce cocoa flavor on roasting. 
* Produces ethanol & heat.
* Metabolic activity stops --> Kills the embryo. 
* Allows the diffusion of metabolites in & out of seed.
* Which triggers complex biochemical reactions in the cotyledon.
:::

## The role of microorganisms during fermentation

* Produce acids and alcohols that penetrate the testae.
* Initiates chemical reactions that will form the precursors.
* No evidence that microbial enzymes penetrate testae.
* Rather enzymes inside the beans are activated by microbial metabolites.
* Exact role of microbes in final product is not understood.


## <font color="#F5F5F5">timeline</font> {data-background-image="images/timeline.png" data-background-size="70%" style="font-size:0em"}


## <font color="#F5F5F5">fermentation process</font> {data-background-image="images/fermentation_process.png" data-background-size="90%" style="font-size:0em"}


## <font color="#F5F5F5">fermented beans</font> { data-background-image="images/fermented_beans.png" data-background-size="85%"  background-transition-speed="slow" style="font-size: 0em;"}


## <font color="#F5F5F5">pre and post testae</font> { data-background-image="images/bean_transform.png" data-background-size="65%"  background-transition-speed="slow" style="font-size: 0em;"}


# Fermentation Dynamics <br/><small>What factors influence community formation & growth kinetics?</small> { style="font-size: 0.8em;"}


## <font color="#F5F5F5">factor</font> { data-background-image="images/factors.png" data-background-size="95%"  background-transition-speed="slow" style="font-size: 0em;"}

# Where do the microbes come from? {style="font-size:0.8em"}

## Do we know?

<font size=6 color="#F9E79F">Seeds</font> within the ripe pod are <font size=6 color="#F9E79F">microbiologically sterile</font>. 

When the <font size=6 color="#F9E79F">pod is opened</font> with a knife, the pulp becomes <font size=6 color="#F9E79F">contaminated</font> with a variety of microorganisms many of which contribute to the subsequent fermentation. 

Organisms come mainly from the hands of workers, knives, unwashed baskets used for transport of seeds, and dried mucilage left on the walls of boxes from previous fermentations.

> <small>from Schwan RF & Wheals AE (2004) The microbiology of cocoa fermentation & its role in chocolate quality. Critical Reviews in Food Science & Nutrition, 44: 205–221</small>

## <font color="#F5F5F5">lambic</font> { data-background-image="images/lambic.png" data-background-size="95%"  background-transition-speed="slow" style="font-size: 0em;"}


# Resources

## <font color="#004949">Photo, image, & graphic credits</font>{data-background="#F5F5F5"}

::: {style="font-size: 0.5em;color: #414141;font-family: 'Lato';font-weight: '100';text-align: left;"}

*Plesse note that unless otherwise stated, I created all figures and  take full responsibility for their content.*

<a href="#/making-cocoa-a-complex-multistep-process">Cacao to cocoa workflow</a> was made by the author using the R package [DiagrammeR](https://rich-iannone.github.io/DiagrammeR/) and adapted from Aprotosoaie, Luca, &  Miron (2015) <em>Flavor Chemistry of Cocoa and Cocoa Products—An Overview.</em> [Comprehensive Reviews in Food Science and Food Safety 15:73-91](https://doi.org/10.1111/1541-4337.12180).

Image  of <a href="#/pre-and-post-testae">pre and post fermented testae</a> was modified from Brillouet & Hue (2017) <em>Fate of proanthocyanidins and anthocyanins along fermentation of cocoa seeds (</em>Theobroma cacao <em>L.</em>). [Journal of Applied Botany and Food Quality 90:141-146.](https://doi.org/10.5073/JABFQ.2017.090.017) Figure 5, page 143. 

Photos on the <a href="#/cacao-to-cocoa">cacao to cocoa slide</a>  credits: left photo entitled *Cocoa beans in a cocoa pod* by Keith Weller ([Kbh3rd](https://en.wikipedia.org/wiki/User:Kbh3rd)) was obtained from Wikimedia Commons licensed under [CC-0](https://creativecommons.org/share-your-work/public-domain/cc0). [Original file](https://commons.wikimedia.org/wiki/File:Cacao-pod-k4636-14.jpg); right photo was obtained from this [website](https://irenamacri.com/ultimate-guide-to-raw-cacao-powder-butter-nibs) (original credit could not be ascertained). 

Close-up photo of  <a href="#/cacao-pulp-a-rich-medium-for-microbial-growth">cacao beans</a>  entitled *bean in pod Cocoa beans in cocoa pod at El Trapiche in Costa Rica* by [Aude](https://commons.wikimedia.org/wiki/User:Aude) was obtained from Wikimedia Commons licensed under [CC BY-SA 3.0](https://creativecommons.org/licenses/by-sa/3.0). [Original file](https://en.wikipedia.org/wiki/Cocoa_bean#/media/File:Cocoa_beans_in_cocoa_pod_at_El_Trapiche,_Costa_Rica.jpg)

Images on the <a href="#/history">History slide </a>come from the following references:

McGovern, Zhang, Tang, Zhang, Hall, Moreau, Nuñez, Butrym, Richards, Wang, Cheng, Zhao, Wang (2004). <em>Fermented beverages of pre- and proto-historic China</em>. [Proceedings of the National Academy of Sciences. 101:17593-17598](https://doi.org/10.1073/pnas.0407921102). Figure 1. 

Henderson, Rosemary, Hall, Gretchen,  Hurst, Jeffrey, McGovern (2007). <em>Chemical and archaeological evidence for the earliest cacao beverages.</em> [Proceedings of the National Academy of Sciences. 104:18937-18940](https://doi.org/10.1073/pnas.0708815104). Figure 2. 

McGovern, Mirzoian, Hall (2009). <em>Ancient Egyptian herbal wines</em>. [Proceedings of the National Academy of Sciences. 106:7361-7366](https://doi.org/10.1073/pnas.0811578106). Figures 1 & 2. 

Boethius (2016). <em>Something rotten in Scandinavia: The world's earliest evidence of fermentation</em>. [Journal of Archaeological Science 66 (2016) 169-180](http://dx.doi.org/10.1016/j.jas.2016.01.008). Figure 10. 

<a href="#/volatile-compound-profiles-in-beans-vs-chocolates">The Sankey diagram </a> was made  using the [Sankey Diagram Generator](http://sankey-diagram-generator.acquireprocure.com/) using data from da Veiga Moreira, de Figueiredo Vilela, Santos, Lima, Schwan (2018). Volatile compounds and protein profiles analyses of fermented cocoa beans and chocolates from different hybrids cultivated in Brazil. [Food research international 109:196-203](https://doi.org/10.1016/j.foodres.2018.04.012). Table 1.

Credits continue on the next slide
:::

## <font color="#004949"> </font>{data-background="#F5F5F5" style="font-size: 1em;"}

::: {style="font-size: 0.5em;color: #414141;font-family: 'Lato';font-weight: '100';text-align: left;"}
:::

<h2 style="text-align: right;">
## <font color="#004949">Creating a chemical collage</font>{style="font-size: 0.9em;font-family: 'Lato';font-weight: '100';" data-background="#F5F5F5" data-vertical-align-top="0px"}
</h2>

::: {style="font-size: 0.7em;color: #414141;font-family: 'Lato';font-weight: '100';text-align: center;"}
Here is the recipe I used to make the <a href="#/volatile-compounds-in-cocoa-flavor">chemical collage</a>. 
:::

#### <font color="#004949">Getting the chemical structures</font>

::: {style="font-size: 0.7em;color: #414141;font-family: 'Lato';font-weight: '100';text-align: left;"}

* Get a list of compound names, e.g., from a paper. 
* Put the names into a single column text file, say `compounds.txt`.
* Upload the file to [PubChem Identifier Exchange Service](https://pubchem.ncbi.nlm.nih.gov/idexchange/idexchange.cgi). Under <u>Input ID List</u> select *Synonyms* from the drop-down menu and the *Choose File* radio button. Upload the file. Under <u>Operator Type</u> select *Same CID* from the drop down. For <u>Output IDs</u>  select *CIDs* and for <u>Output Method</u> choose the *two column* option. Then hit <u>Submit Job</u>. 
* Download the resulting file. This will have the CID to Compound name. Delete the compound name so there is a single column of CIDs. Save a new file with the `csv` file extension. 
* Upload file to the [PubChem Download Service](https://pubchem.ncbi.nlm.nih.gov/pc_fetch/pc_fetch.cgi). Select <u>PubChem Compound</u>, under <u>Choose a format</u> select *Image* then hit the <u>Download</u> button. You should get a directory of `.png` image files, one for each CID.
:::

<br/>

#### <font color="#004949">Putting the pieces together</font>

::: {style="font-size: 0.7em;color: #414141;font-family: 'Lato';font-weight: '100';text-align: left;"}
* Once I had `.png` files for all of the compounds I did some cropping in [Gimp](https://www.gimp.org/).
* Then I used an online tool called [Photo Collage](https://www.photocollage.com/) to put all the images together and [Inkscape](https://inkscape.org/) to add text and finalize the image. Finally I  used  [this tool](https://www.websiteplanet.com/webtools/imagecompressor/) to compress the image. 
::: 


## Packages used in this presentation

::: {style="font-size: 0.8em;font-family: 'Lato';font-weight: '100';"}

*  [reveal.js](https://github.com/hakimel/reveal.js) An HTML Presentation Framework in [R Markdown](https://bookdown.org/yihui/rmarkdown/revealjs.html).  
* [DiagrammeR](https://rich-iannone.github.io/DiagrammeR/) an R package for flow charts.  
*  [treemap](https://github.com/mtennekes/treemap) an R package for treemap visualization.  

:::


